---
title: "218X Assignment 6"
author: "Lena Bakalian"
date: "10/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = F, message = F)
```


```{r}
library(censusapi)
library(tidyverse)
library(tigris)
library(sf)
library(leaflet)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")

pums_2019_1yr <- getCensus(
  name = "acs/acs1/pums",
  vintage = 2019,
  region = "public use microdata area:*",
  regionin = "state:06",
  vars = c(
    "SERIALNO",
    "SPORDER",
    "PWGTP",
    "WGTP",
    "YBL",
    "BLD",
    "MV",
    "HINCP",
    "TEN",
    "AGEP"
  )
)

saveRDS(pums_2019_1yr, "a6_pums.rds")
pums_2019_1yr <- readRDS("a6_pums.rds")
```

```{r}
#First, filter to San Francisco PUMAs.
ca_pumas <-
  pumas("CA", cb = T, progress_bar = F)

sf_county <-
  counties("CA", cb = T, progress_bar = F) %>% 
  filter(NAME == "San Francisco") 

sf_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[sf_county, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()


sf_pums <-
  pums_2019_1yr %>% 
  mutate(
    PUMA = str_pad(public_use_microdata_area,5,"left","0")
  ) %>% 
  filter(PUMA %in% sf_pumas$PUMACE10)
```


```{r}
#Filter to housing structures built before 1960. This is a simplification of the more complicated breakdown shown in Table 2 of the Appendix of the Lead Risk Chapter in the CES report.

sf_pums_clean <- sf_pums %>% 
  mutate(
    YBL = as.numeric(YBL),
    AGEP = as.numeric(AGEP),
  ) %>% 
  filter(as.numeric(YBL) %in% 1:3) %>% 
#OR: filter(YBL %in% as.character(1:3))
  arrange(AGEP) %>% 
  group_by(SERIALNO) %>% 
  summarize_all(first)


#Use mutate() to create a variable called leadrisk, which can be set to 1 when two conditions are both true (use & to link two conditional statements about HINCP and AGEP).

sf_pums_regression <-
  sf_pums_clean %>%  
  mutate(
    leadrisk = ifelse(
    as.numeric(HINCP) <= 90000 & AGEP <= 6,
    1,
    0
  )
)
```


```{r}
#Create a logit model in which leadrisk is the predicted variable, and BLD, TEN, MV, and PUMA are the predictors. 

sf_pums_factor <- sf_pums_regression %>% 
  mutate(
    unit = BLD %>% 
      factor(
        levels = sf_pums$BLD %>% 
          unique() %>% 
          as.numeric() %>% 
          sort()
       ),
      tenure = TEN %>% 
        factor(
        levels = sf_pums$TEN %>% 
          unique() %>% 
          as.numeric() %>% 
          sort()
      ),
      time = MV %>% 
        factor(
        levels = sf_pums$MV %>% 
          unique() %>% 
          as.numeric() %>% 
          sort()
      ),
      puma = PUMA %>% 
        factor(
        levels = sf_pums$PUMA %>% 
          unique() %>% 
          sort()
      ))
```


```{r}
#Show a summary of the model
logit_model <- glm(
  leadrisk ~ unit + tenure + time + puma,
  family = quasibinomial(),
  data = sf_pums_factor
)

summary(logit_model)

```


```{r}
#detail one example of a predicted likelihood score (pick a random record, state the predictor variables, and state the predicted score).

#grab one random row from your full dataset, using sample_n() (look this function up to see the simple way to use it). Then just supply this one-row dataframe inside of predict().

sample_predict <- sample_n(sf_pums_factor, 1)

predict(logit_model, sample_predict, type = "response")
```

```{r}
#Calculate the predicted score (between 0 and 1) for your entire dataset. 

sf_pums_predicted <- sf_pums_factor %>% 
  mutate(predict = predict(logit_model, sf_pums_factor, type = "response"))

#Present true positives, true negatives, and Type 1/2 errors of your model’s performance on identifying actual “lead risk” (income < $90K, at least one child) in these households, where your threshold score is 0.10 or more (10%). 

summary_2x2 <-
  sf_pums_predicted %>% 
  mutate(
    leadrisk = ifelse(
      leadrisk == 1, 
      "Yes (At risk)", 
      "No (Not at risk)"
    ),
    predicted_yes = ifelse(
      predict >= 0.1, 
      as.numeric(WGTP),
      0
    ),
      predicted_no = ifelse(
      predict <= 0.1, 
      as.numeric(WGTP),
      0
      
  )) %>% 
group_by(leadrisk) %>% 
summarize(
  `Predicted_Yes` = sum(predicted_yes), 
  `Predicted_No` = sum(predicted_no)
)

summary_2x2
```
Analysis:

The households which exceeded the threshold score (leadrisk > 1), BLD values were generally referencing apartments or attached single family homes. For tenure (TEN values), many of the values were tied to residences either owned with a mortgage or loan or rented. For MV values, many reflected that the residents moved into the house or apartment in the range of 5 to 9 years ago. 

Adding the true and false positives, you'd get 5,368 at risk. Therefore you'd have this many households to mail postcards to. 

I'd expect 16.2% of postcards to actually go to low-income households with children (true positive/all positive). These are the true positive households that we have accounted for in our model. Therefore, only 16.2% of all at-risk children are likely to be reached. 86% of children reached (false negative/total predicted yes) are not at risk.

Ways to improve strategy might be to target schools/preschools and deliver postcards there. You might deliver them to students who are not at risk, but you will likely reach a higher percentage of children who are at risk. Another way could be to have them at doctors offices for parents to take. 
 
