---
title: "CE218X Assignment 2"
author: "Lena Bakalian"
date: "9/30/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = F, message = F)
```

```{r}
#loading libraries
library(tidyverse)
library(sf)
library(censusapi)
library(mapview)
library(leaflet)
library(devtools)
library(tigris)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")
```

```{r}
#population data 2020
dec_vars_2020 <-
  listCensusMetadata(
    name = "2020/dec/pl",
    type = "variables"
  )
scc_pop_2020 <- getCensus(
  name = "dec/pl",
  vintage = 2020,
  region = "block:*", 
  regionin = "state:06+county:085",
  vars = "P1_006N"
) %>% 
  transmute(
    GEOID20 = paste0(state,county,tract,block),
    pop_2020 = P1_006N
  )
```   


```{r}
#population data 2010
dec_vars_2010 <-
  listCensusMetadata(    
    name = "2010/dec/pl",
    type = "variables")
scc_pop_2010 <-
  getCensus(
  name = "dec/pl",
  vintage = 2010,
  region = "block:*", 
  regionin = "state:06+county:085",
  vars = "P001006"
  ) %>% 
  transmute(
    GEOID10 = paste0(state,county,tract,block),
    pop_2010 = P001006
  )
```

#constructing the aligned geometry set for blocks 2010
```{r}
blocks_2020 <- blocks("CA", "Santa Clara", year = 2020, progress_bar = F)
blocks_2010 <- blocks("CA", "Santa Clara", year = 2010, progress_bar = F)

place_boundary <- places("CA", progress_bar = F) %>% 
  filter(NAME %in% "Cupertino")

place_blocks_2010 <- scc_pop_2010 %>% 
  left_join(blocks_2010 %>% select(GEOID10)) %>% 
  st_as_sf() %>% 
  st_centroid() %>% 
  .[place_boundary, ] %>% 
  st_set_geometry(NULL) %>% 
  left_join(blocks_2010 %>% select(GEOID10)) %>%
  st_as_sf()
  
mapview(place_boundary)+ mapview(place_blocks_2010, alpha.regions = 0, color = "red", lwd = 2)

```


```{r}
#2020 blocks that intersect with 2010 blocks (GEOID)
place_blocks_2020 <- blocks_2020 %>%
  .[place_blocks_2010, ] %>% 
  select(GEOID20) %>% 
left_join(scc_pop_2020)
```


```{r}
#spatial subsetting 
place_blocks_2020_intersect <- place_blocks_2020 %>% 
  st_transform(26910) %>% 
  mutate(original_area = st_area(.)) %>% 
  st_intersection(
    place_blocks_2010 %>%
      select(GEOID10) %>% 
      st_transform(26910)) %>% 
  
  mutate(
    leftover_area = st_area(.), 
    perc_area = as.numeric(leftover_area / original_area), 
    pop_new = pop_2020 * perc_area)
```


```{r}
#join back to the 2010 data
place_blocks_2020_reshaped <- place_blocks_2020_intersect %>%
  st_drop_geometry() %>% 
  group_by(GEOID10) %>% 
  summarize(
    pop_2020 = sum(pop_new, na.rm=T) %>% round()
  )
```


```{r}
#join this to place_blocks_2010 and compute the absolute difference
projection <- "+proj=utm +zone=10 +ellps=GRS80 +datum=NAD83 +units=ft +no_defs"
place_pop_2010_2020 <- place_blocks_2010 %>% 
  left_join(place_blocks_2020_reshaped) %>% 
  mutate(
    absolute_pop = (pop_2020 - pop_2010) / as.numeric(st_area(.)) * 43560
  )

```
```{r}
max_variable <- max(abs(place_pop_2010_2020$absolute_pop))
absolute_pal <- colorNumeric(
  palette = "RdBu",
  domain = c(-max_variable,max_variable)
)
```


```{r}
#map
leaflet() %>% 
  addProviderTiles(provider = providers$CartoDB.Positron) %>%
  addPolygons(
    data = place_pop_2010_2020,
    fillColor = ~absolute_pal(absolute_pop),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.75,
    weight = 1,
    label = ~round(absolute_pop),
  ) %>% 
  addLegend(
    pal = absolute_pal,
    value = c(-max_variable,max_variable),
    title = "Asian population in Cupertino, CA from 2010 to 2020"
)
```

Reflections: In the analysis, I assumed that the data I was using accurately reflected population changes, since the data was self reported. My assumption was that 2020 was the constraining year, so I used this data to "cut" the 2010 data. I looked at population of Asian people in Cupertino. After plotting this data, it seems as though there was not too much of a significant change in this population in this area. Since this area has had more growth in recent years, it surprised me a bit that there was not more of an increase in population here. 

